Safe Redis Queries

### 2016-06-11 aplas16/sections/TypeLevelDict.tex
Tweaks in Section 2.

One of the challenges of statically ensuring type correctness of \Redis{},
which also presents in other stateful languages, is that the type of the value
associated to a key can be altered after updating. To ensure type correctness,
we have to keep track of the (\Redis{}) types of all existing keys in a
{\em dictionary} --- conceptually, a list of pairs of keys and \Redis{} types.
Each \Redis{} command is embedded in \Popcorn{} as a monadic computation. The
monad, to be presented in Section \ref{sec:indexed-monads}, is indexed by
the dictionaries before and after the computation. In a dependently typed
programming language (without the so-called ``phase distinction'' ---
separation between types and terms), this would pose no problem. In Haskell
however, the dictionaries, to index a monad, has to be a Haskell type as well.

In this section we describe how to construct a type-level dictionary, to be
used with the indexed monad in Section \ref{sec:indexed-monads}. More operations
on the dictionary will be presented in Section \ref{sec:type-level-fun}.

@@

Normally, at the term level, we could express the datatype of dictionary with
\emph{type synonym} like this.\footnotemark

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathbf{type}\;\Conid{Key}\mathrel{=}\Conid{String}{}\<[E]%
\\
\>[B]{}\mathbf{type}\;\Conid{Dictionary}\mathrel{=}[\mskip1.5mu (\Conid{Key},\Conid{TypeRep})\mskip1.5mu]{}\<[E]%
\ColumnHook
\end{hscode}\resethooks

\footnotetext{\ensuremath{\Conid{TypeRep}} supports term-level representations
 of datatypes, available in \ensuremath{\Conid{\Conid{Data}.Typeable}}}

To encode this in the type level, everything has to be
 \emph{promoted}\cite{promotion} one level up.
 From terms to types, and from types to kinds.

Luckily, with recently added GHC extension \emph{data kinds}, suitable
 datatype will be automatically promoted to be a kind, and its value
 constructors to be type constructors. The following type \ensuremath{\Conid{List}}

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathbf{data}\;\Conid{List}\;\Varid{a}\mathrel{=}\Conid{Nil}\mid \thinspace'\Varid{Cons}\;\Varid{a}\;(\Conid{List}\;\Varid{a}){}\<[E]%
\ColumnHook
\end{hscode}\resethooks

Give rise to the following kinds and type constructors:\footnote{To distinguish
 between types and promoted constructors that have
 ambiguous names, prefix promoted constructor with a single quote like
 \ensuremath{\thinspace'\Varid{Nil}} and \ensuremath{\Conid{CONS}}}
\footnote{All kinds have \emph{sort} BOX in Haskell\cite{sorts}}

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{6}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Conid{List}\;\Varid{k}\mathbin{::}\Conid{BOX}{}\<[E]%
\\
\>[B]{}\Conid{Nil}{}\<[6]%
\>[6]{}\mathbin{::}\Conid{List}\;\Varid{k}{}\<[E]%
\\
\>[B]{}\thinspace'\Varid{Cons}\mathbin{::}\Varid{k}\to \Conid{List}\;\Varid{k}\to \Conid{List}\;\Varid{k}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks

Haskell sugars lists \ensuremath{[\mskip1.5mu \mathrm{1},\mathrm{2},\mathrm{3}\mskip1.5mu]} and tuples
 \ensuremath{(\mathrm{1},\text{\tt 'a'})} with brackets and parentheses.
 We could also express promoted lists and tuples in types like this with
 a single quote prefixed. For example:
 \ensuremath{\thinspace'[\!\;\Conid{Int},\Conid{Char}\;\!]}, \ensuremath{\thinspace'(\!\;\Conid{Int},\Conid{Char}\;\!)}.

>>>>

One of the challenges of statically ensuring type correctness of \Redis{},
which also presents in other stateful languages, is that the type of the value
associated to a key can be altered after updating. To ensure type correctness,
we have to keep track of the (\Redis{}) types of all existing keys in a
{\em dictionary} --- conceptually, a list of pairs of keys and \Redis{} types.
Each \Redis{} command is embedded in \Popcorn{} as a monadic computation. The
monad, to be presented in Section~\ref{sec:indexed-monads}, is indexed by
the dictionaries before and after the computation. In a dependently typed
programming language (without the so-called ``phase distinction'' ---
separation between types and terms), this would pose no problem. In Haskell
however, the dictionaries, to index a monad, has to be a Haskell type as well.

In this section we describe how to construct a type-level dictionary, to be
used with the indexed monad in Section~\ref{sec:indexed-monads}. More operations
on the dictionary will be presented in Section~\ref{sec:type-level-fun}.

@@

% Normally, at the term level, we could express the datatype of dictionary with
% \emph{type synonym} like this.\footnotemark
%
% \begin{spec}
% type Key = String
% type Dictionary = [(Key, TypeRep)]
% \end{spec}
%
% \footnotetext{|TypeRep| supports term-level representations
%  of datatypes, available in |Data.Typeable|}

A datatype definition such as the one below:
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathbf{data}\;\Conid{List}\;\Varid{a}\mathrel{=}\Conid{Nil}\mid \Conid{Cons}\;\Varid{a}\;(\Conid{List}\;\Varid{a})~~,{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
is usually understood as having defined a type constructor \ensuremath{\Conid{List}}, and two value
constructors \ensuremath{\Conid{Nil}} and \ensuremath{\Conid{Cons}}. For all lifted types \ensuremath{\Varid{a}}, \ensuremath{\Conid{List}\;\Varid{a}} is also a
lifted type. Thus the {\em kind} of \ensuremath{\Conid{List}} is \ensuremath{\mathbin{*}\to \mathbin{*}}, where \ensuremath{\mathbin{*}} is the kind of
all {\em lifted types} in Haskell. The two value constructors respectively
have types \ensuremath{\Conid{Nil}\mathbin{::}\Conid{List}\;\Varid{a}} and \ensuremath{\Conid{Cons}\mathbin{::}\Varid{a}\to \Conid{List}\;\Varid{a}\to \Conid{List}\;\Varid{a}}.

The GHC extension \emph{data kinds}~\cite{promotion}, however, automatically
promotes certain ``suitable'' types to kinds. With the extension, the \ensuremath{\mathbf{data}}
definition has an alternative reading: for all kind \ensuremath{\Varid{k}}, \ensuremath{\Conid{List}\;\Varid{k}} is also a
kind. \ensuremath{\Conid{Nil}} is a type, whose kind is \ensuremath{\Conid{List}\;\Varid{k}} for some \ensuremath{\Varid{k}}. Given a type \ensuremath{\Varid{x}}
of kind \ensuremath{\Varid{k}} and a type \ensuremath{\Varid{xs}} of kind \ensuremath{\Conid{List}\;\Varid{k}}, \ensuremath{\Conid{Cons}\;\Varid{x}\;\Varid{xs}} is again a type of
kind \ensuremath{\Conid{List}\;\Varid{k}}. Formally, for kind \ensuremath{\Varid{k}}, \ensuremath{\Conid{Nil}\mathbin{::}\Conid{List}\;\Varid{k}} and \ensuremath{\Conid{Cons}\mathbin{::}\Varid{k}\to \Conid{List}\;\Varid{k}\to \Conid{List}\;\Varid{k}}. Whether a constructor is promoted can often be inferred
from the context. To be more specific, prefixing a contructor with
a single quote, such as in \ensuremath{\thinspace'\Conid{Nil}} and \ensuremath{\thinspace'\Conid{Cons}}, denotes that it is promoted.

The built-in lists in Haskell has two constructors \ensuremath{[\mskip1.5mu \mskip1.5mu]} and \ensuremath{(\mathbin{:})}, and
\ensuremath{[\mskip1.5mu \mathrm{1},\mathrm{2},\mathrm{3}\mskip1.5mu]} is take as a syntax sugar for \ensuremath{\mathrm{1}\mathbin{:}(\mathrm{2}\mathbin{:}(\mathrm{3}\mathbin{:}[\mskip1.5mu \mskip1.5mu]))}. The type and
data constructors can also be promoted. Given kind \ensuremath{\Varid{k}}, \ensuremath{[\mskip1.5mu \Varid{k}\mskip1.5mu]} is also a kind.
The type constructor 
For example, since \ensuremath{\Conid{Int}}, \ensuremath{\Conid{Char}}, etc., all have kind \ensuremath{\mathbin{*}}, \ensuremath{\Conid{Int}\mathbin{\thinspace':}(\Conid{Char}\mathbin{\thinspace':}(\Conid{Bool}\mathbin{\thinspace':}\thinspace'[]))} is a type having kind \ensuremath{[\mskip1.5mu \mathbin{*}\mskip1.5mu]} -- a list of types. The same list can be denoted by a syntax
sugar \ensuremath{\thinspace'[\!\;\Conid{Int},\Conid{Char},\Conid{Bool}\;\!]}.

sugar for \ensuremath{\mathrm{1}\mathbin{\thinspace':}(\mathrm{2}\mathbin{\thinspace':}(\mathrm{3}\mathbin{\thinspace':}\thinspace'[]))}
Haskell sugars lists \ensuremath{[\mskip1.5mu \mathrm{1},\mathrm{2},\mathrm{3}\mskip1.5mu]} and tuples
 \ensuremath{(\mathrm{1},\text{\tt 'a'})} with brackets and parentheses.
 We could also express promoted lists and tuples in types like this with
 a single quote prefixed. For example:
 \ensuremath{\thinspace'[\!\;\Conid{Int},\Conid{Char}\;\!]}, \ensuremath{\thinspace'(\!\;\Conid{Int},\Conid{Char}\;\!)}.
